---
title: "TMRC2 `r Sys.getenv('VERSION')`: Macrophage Differential Expression."
author: "atb abelew@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
---

<style type="text/css">
body .main-container {
  max-width: 1600px;
}
body, td {
  font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
 font-size: 16px
}
</style>

```{r options, include = FALSE}
library(ggplot2)
library(glue)
library(Heatplus)
library(hpgltools)
library(UpSetR)
library(tibble)

knitr::opts_knit$set(progress = TRUE, verbose = TRUE, width = 90, echo = TRUE)
knitr::opts_chunk$set(
  error = TRUE, fig.width = 8, fig.height = 8, fig.retina = 2,
  out.width = "100%", dev = "png",
  dev.args = list(png = list(type = "cairo-png")))
old_options <- options(digits = 4, stringsAsFactors = FALSE, knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
ver <- Sys.getenv("VERSION")
previous_file <- ""
rundate <- format(Sys.Date(), format = "%Y%m%d")

## tmp <- try(sm(loadme(filename = gsub(pattern = "\\.Rmd", replace = "\\.rda\\.xz", x = previous_file))))
rmd_file <- "03differential_expression.Rmd"
loaded <- load(file = glue("rda/tmrc2_data_structures-v{ver}.rda"))
savefile <- gsub(pattern = "\\.Rmd", replace = "\\.rda\\.xz", x = rmd_file)
```

# Changelog

* 202310: Cleaning up to make everything pass within a containerized environment.
* 202310: Received a set of colors and contrasts of interest for a barplot of significance.
* 20230410: Making some changes to improve the differential expression
plots as well as prepare for some different pathway/GSEA/GSVA
analyses on the data.

# Notes

Showing plots of counts with respect to drug treatment: query to Najib
to clarify normalization. I think I can make versions of these plots
which use SL/normalized counts to alleviate Najib's query.

mean ratio of SL/total by condition?

# Introduction

Having established that the TMRC2 macrophage data looks robust and
illustrative of a couple of interesting questions, let us perform a
couple of differential analyses of it.

Also note that as of 202212, we received a new set of samples which
now include some which are of a completely different cell type,
U937. As their ATCC page states, they are malignant cells taken from
the pleural effusion of a 37 year old white male with histiocytic
lymphoma and which exhibit the morphology of monocytes.  Thus, this
document now includes some comparisons of the cell types as well as
the various macrophage donors (given that there are now more donors
too).

## Human data

I am moving the dataset manipulations here so that I can look at them
all together before running the various DE analyses.

## Create sets focused on drug, celltype, strain, and combinations

Let us start by playing with the metadata a little and create sets
with the condition set to:

* Drug treatment
* Cell type (macrophage or U937)
* Donor
* Infection Strain
* Some useful combinations thereof

In addition, keep mental track of which datasets are comprised of all
samples vs. those which are only macrophage vs. those which are only
U937.  (Thus, the usage of all_human vs. hs_macr vs. u937 as prefixes
for the data structures.)

Ideally, these recreations of the data should perhaps be in the
datastructures worksheet.

```{r}
all_human <- sanitize_expt_pData(hs_macrophage, columns = "drug") %>%
  set_expt_conditions(fact = "drug") %>%
  set_expt_batches(fact = "typeofcells")

## The following 3 lines were copy/pasted to datastructures and should be removed soon.
no_strain_idx <- pData(all_human)[["strainid"]] == "none"
##pData(all_human)[["strainid"]] <- paste0("s", pData(all_human)[["strainid"]],
##                                         "_", pData(all_human)[["macrophagezymodeme"]])
pData(all_human)[no_strain_idx, "strainid"] <- "none"
table(pData(all_human)[["strainid"]])

all_human_types <- set_expt_conditions(all_human, fact = "typeofcells") %>%
  set_expt_batches(fact = "drug")

type_zymo_fact <- paste0(pData(all_human_types)[["condition"]], "_",
                         pData(all_human_types)[["macrophagezymodeme"]])
type_zymo <- set_expt_conditions(all_human_types, fact = type_zymo_fact)

type_drug_fact <- paste0(pData(all_human_types)[["condition"]], "_",
                         pData(all_human_types)[["drug"]])
type_drug <- set_expt_conditions(all_human_types, fact = type_drug_fact)

strain_fact <- pData(all_human_types)[["strainid"]]
table(strain_fact)

new_conditions <- paste0(pData(hs_macrophage)[["macrophagetreatment"]], "_",
                         pData(hs_macrophage)[["macrophagezymodeme"]])
## Note the sanitize() call is redundant with the addition of sanitize() in the
## datastructures file, but I don't want to wait to rerun that.
hs_macr <- set_expt_conditions(hs_macrophage, fact = new_conditions) %>%
  sanitize_expt_pData(column = "drug") %>%
  subset_expt(subset = "typeofcells!='U937'")
```

### Separate Macrophage samples

Once again, we should reconsider where the following block is placed,
but these datastructures are likely to be used in many of the
following analyses.

```{r}
hs_macr_drug_expt <- set_expt_conditions(hs_macr, fact = "drug")

hs_macr_strain_expt <- set_expt_conditions(hs_macr, fact = "macrophagezymodeme") %>%
  subset_expt(subset = "macrophagezymodeme != 'none'")

table(pData(hs_macr)[["strainid"]])
```

### Refactor U937 samples

The U937 samples were separated in the datastructures file, but we
want to use the combination of drug/zymodeme with them pretty much
exclusively.

```{r}
new_conditions <- paste0(pData(hs_u937)[["macrophagetreatment"]], "_",
                         pData(hs_u937)[["macrophagezymodeme"]])
u937_expt <- set_expt_conditions(hs_u937, fact = new_conditions)
```

## Contrasts used in this document

Given the various ways we have chopped up this dataset, there are a
few general types of contrasts we will perform, which will then be
combined into greater complexity:

* drug treatment
* strains used
* cellltypes
* donors

In the end, our actual goal is to consider the variable effects of
drug+strain and see if we can discern patterns which lead to better or
worse drug treatment outcome.

There is a set of contrasts in which we are primarily interested in
this data, these follow.  I created one ratio of ratios contrast which
I think has the potential to ask our biggest question.

```{r}
tmrc2_human_extra <- "z23drugnodrug_vs_z22drugnodrug = (infsbz23 - infz23) - (infsbz22 - infz22), z23z22drug_vs_z23z22nodrug = (infsbz23 - infsbz22) - (infz23 - infz22)"
tmrc2_human_keepers <- list(
  "z23nosb_vs_uninf" = c("infz23", "uninfnone"),
  "z22nosb_vs_uninf" = c("infz22", "uninfnone"),
  "z23nosb_vs_z22nosb" = c("infz23", "infz22"),
  "z23sb_vs_z22sb" = c("infsbz23", "infsbz22"),
  "z23sb_vs_z23nosb" = c("infsbz23", "infz23"),
  "z22sb_vs_z22nosb" = c("infsbz22", "infz22"),
  "z23sb_vs_sb" = c("infsbz23", "uninfsbnone"),
  "z22sb_vs_sb" = c("infsbz22", "uninfsbnone"),
  "z23sb_vs_uninf" = c("infsbz23", "uninfnone"),
  "z22sb_vs_uninf" = c("infsbz22", "uninfnone"),
  "sb_vs_uninf" = c("uninfsbnone", "uninfnone"),
  "extra_z2322" = c("z23drugnodrug", "z22drugnodrug"),
  "extra_drugnodrug" = c("z23z22drug", "z23z22nodrug"))
single_tmrc2_keeper <- list(
  "z22sb_vs_sb" = c("infsbz22", "uninfsbnone"))
tmrc2_drug_keepers <- list(
  "drug" = c("antimony", "none"))
tmrc2_type_keepers <- list(
  "type" = c("U937", "Macrophages"))
tmrc2_strain_keepers <- list(
  "strain" = c("z23", "z22"))
type_zymo_extra <- "zymos_vs_types = (U937z23 - U937z22) - (Macrophagesz23 - Macrophagesz22)"
tmrc2_typezymo_keepers <- list(
  "u937_macr" = c("Macrophagesnone", "U937none"),
  "zymo_macr" = c("Macrophagesz23", "Macrophagesz22"),
  "zymo_u937" = c("U937z23", "U937z22"),
  "z23_types" = c("U937z23", "Macrophagesz23"),
  "z22_types" = c("U937z22", "Macrophagesz22"),
  "zymos_types" = c("zymos_vs_types"))
tmrc2_typedrug_keepers <- list(
  "type_nodrug" = c("U937none", "Macrophagesnone"),
  "type_drug" = c("U937antimony", "Macrophagesantimony"),
  "macr_drugs" = c("Macrophagesantimony", "Macrophagesnone"),
  "u937_drugs" = c("U937antimony", "U937none"))
u937_keepers <- list(
  "z23nosb_vs_uninf" = c("infz23", "uninfnone"),
  "z22nosb_vs_uninf" = c("infz22", "uninfnone"),
  "z23nosb_vs_z22nosb" = c("infz23", "infz22"),
  "z23sb_vs_z22sb" = c("infsbz23", "infsbz22"),
  "z23sb_vs_z23nosb" = c("infsbz23", "infz23"),
  "z22sb_vs_z22nosb" = c("infsbz22", "infz22"),
  "z23sb_vs_sb" = c("infsbz23", "uninfsbnone"),
  "z22sb_vs_sb" = c("infsbz22", "uninfsbnone"),
  "z23sb_vs_uninf" = c("infsbz23", "uninfnone"),
  "z22sb_vs_uninf" = c("infsbz22", "uninfnone"),
  "sb_vs_uninf" = c("uninfsbnone", "uninfnone"))
high_expression <- 128
high_expression_column <- "deseq_basemean"

combined_to_tsv <- function(combined, celltype = "all") {
  keepers <- combined[["keepers"]]
  for (k in seq_len(length(keepers))) {
    kname <- names(keepers)[k]
    numerator <- keepers[[k]][1]
    denominator <- keepers[[k]][2]
    filename <- glue("excel/macrophage_de/{ver}/tsv_tables/tmrc2_{celltype}_{kname}_n{numerator}_d{denominator}-v{ver}.xlsx")
    kdata <- combined[["data"]][[kname]]
    if (is.null(kdata[["basic_num"]])) {
      next
    }
    wanted <- c("hgncsymbol", "deseq_logfc", "deseq_adjp", "deseq_basemean", "deseq_num", "deseq_den")
    wanted_data <- kdata[, wanted]
    colnames(wanted_data) <- c("hgncsymbol", "deseq_logfc", "deseq_adjp", "deseq_mean", "deseq_numerator", "deseq_denominator")
    write_xlsx(data = wanted_data, excel = filename)
  }
}

write_all_gp <- function(all_gp) {
  for (g in seq_len(length(all_gp))) {
    name <- names(all_gp)[g]
    datum <- all_gp[[name]]
    filename <- glue("excel/macrophage_de/{ver}/gprofiler/{name}_gprofiler-v{ver}.xlsx")
    written <- sm(write_gprofiler_data(datum, excel = filename))
  }
}
```

### Primary queries

There is a series of initial questions which make some sense
to me, but these do not necessarily match the set of questions which
are most pressing.  I am hoping to pull both of these sets of
queries in one.

Before extracting these groups of queries, let us invoke the
all_pairwise() function and get all of the likely contrasts along with
one or more extras that might prove useful (the 'extra' argument).

### Combined U937 and Macrophages: Compare drug effects

When we have the u937 cells in the same dataset as the macrophages,
that provides an interesting opportunity to see if we can observe
drug-dependant effects which are shared across both cell types.

```{r}
drug_de <- all_pairwise(all_human, filter = TRUE, model_batch = "svaseq", do_noiseq = FALSE)
drug_de

drug_table <- combine_de_tables(
  drug_de, keepers = tmrc2_drug_keepers,
  excel = glue("excel/macrophage_de/{ver}/de_tables/macrophage_drug_comparison-v{ver}.xlsx"))
drug_table
combined_to_tsv(drug_table, celltype = "all")

drug_sig <- extract_significant_genes(
  drug_table,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_drug_sig-v{ver}.xlsx"))
drug_sig
drug_highsig <- extract_significant_genes(
  drug_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_drug_highsig-v{ver}.xlsx"))
drug_highsig
```

```{r}
all_drug_gp <- all_gprofiler(drug_sig)
all_drug_gp
write_all_gp(all_drug_gp)
```

### Combined U937 and Macrophages: compare cell types

There are a couple of ways one might want to directly compare the two
cell types.

* Given that the variance between the two celltypes is so huge, just
compare all samples.
* One might want to compare them with the interaction effects of drug/zymodeme.

```{r}
type_de <- all_pairwise(all_human_types, filter = TRUE, model_batch = "svaseq", do_noiseq = FALSE)
type_de

type_table <- combine_de_tables(
  type_de, keepers = tmrc2_type_keepers,
  excel = glue("excel/macrophage_de/{ver}/de_tables/macrophage_type_comparison-v{ver}.xlsx"))
type_table
combined_to_tsv(type_table, celltype = "all")

type_sig <- extract_significant_genes(
  type_table,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_sig-v{ver}.xlsx"))
type_sig
type_highsig <- extract_significant_genes(
  type_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_highsig-v{ver}.xlsx"))
type_highsig
```

#### Combined factors of interest: celltype+zymodeme

Given the above explicit comparison of all samples comprising the two
cell types, now let us look at the drug treatment+zymodeme status with
all samples, macrophages and U937.

```{r}
type_zymo_de <- all_pairwise(type_zymo, filter = TRUE, model_batch = "svaseq", do_noiseq = FALSE,
                             extra_contrasts = type_zymo_extra)
type_zymo_de

type_zymo_table <- combine_de_tables(
  type_zymo_de, keepers = tmrc2_typezymo_keepers,
  excel = glue("excel/macrophage_de/de_tables/macrophage_type_zymo_comparison-v{ver}.xlsx"))
combined_to_tsv(type_zymo_table, celltype = "all")

type_zymo_sig <- extract_significant_genes(
  type_zymo_table,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_zymo_sig-v{ver}.xlsx"))
type_zymo_sig
type_zymo_highsig <- extract_significant_genes(
  type_zymo_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_zymo_highsig-v{ver}.xlsx"))
```

#### Combined factors of inteest: celltype+drug

The 'type_drug' datastructure is the same as above, but the condition
is created from the concatenation of the cell type and drug treatment.

```{r}
type_drug_de <- all_pairwise(type_drug, filter = TRUE, model_batch = "svaseq")
type_drug_de
type_drug_table <- combine_de_tables(
  type_drug_de, keepers = tmrc2_typedrug_keepers,
  excel = glue("excel/macrophage_de/{ver}/de_tables/macrophage_type_drug_comparison-v{ver}.xlsx"))
type_drug_table
combined_to_tsv(type_drug_table, celltype = "all")
type_drug_sig <- extract_significant_genes(
  type_drug_table,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_drug_sig-v{ver}.xlsx"))
type_drug_sig
type_drug_highsig <- extract_significant_genes(
  type_drug_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/{ver}/sig_tables/macrophage_type_drug_highsig-v{ver}.xlsx"))
type_drug_highsig
```

# Individual cell types

At this point, I think it is fair to say that the two cell types are
sufficiently different that they do not really belong together in a
single analysis.

## drug or strain effects, single cell type

One of the queries Najib asked which I think I misinterpreted was to
look at drug and/or strain effects.  My interpretation is somewhere
below and was not what he was looking for.  Instead, he was looking to
see all(macrophage) drug/nodrug and all(macrophage) z23/z22 and
compare them to each other.  It may be that this is still a wrong
interpretation, if so the most likely comparison is either:

*  (z23drug/z22drug) / (z23nodrug/z22nodrug), or perhaps
*  (z23drug/z23nodrug) / (z22drug/z22nodrug),

I am not sure those confuse me, and at least one of them is below

### Macrophages

In these blocks we will explicitly query only one factor at a time,
drug and strain.  The eventual goal is to look for effects of
drug treatment and/or strain treatment which are shared?

#### Macrophage Drug only

Thus we will start with the pure drug query.  In this block we will
look only at the drug/nodrug effect.

```{r}
hs_macr_drug_de <- all_pairwise(hs_macr_drug_expt, filter = TRUE, model_batch = "svaseq")
hs_macr_drug_de
hs_macr_drug_table <- combine_de_tables(
  hs_macr_drug_de, keepers = tmrc2_drug_keepers,
  excel = glue("excel/macrophage_de/macrophage_onlydrug_table-v{ver}.xlsx"))
hs_macr_drug_table
combined_to_tsv(hs_macr_drug_table, celltype = "macrophage")

hs_macr_drug_sig <- extract_significant_genes(
  hs_macr_drug_table,
  excel = glue("excel/macrophage_de/macrophageonly_drug_sig-v{ver}.xlsx"))
hs_macr_drug_sig
hs_macr_drug_highsig <- extract_significant_genes(
  hs_macr_drug_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/macrophageonly_drug_highsig-v{ver}.xlsx"))
hs_macr_drug_highsig
```

#### Macrophage Strain only

In a similar fashion, let us look for effects which are observed when
we consider only the strain used during infection.

```{r}
hs_macr_strain_de <- all_pairwise(hs_macr_strain_expt, filter = TRUE, model_batch = "svaseq")
hs_macr_strain_de
hs_macr_strain_table <- combine_de_tables(
  hs_macr_strain_de, keepers = tmrc2_strain_keepers,
  excel = glue("excel/macrophage_de/macrophage_onlystrain_table-v{ver}.xlsx"))
hs_macr_strain_table
combined_to_tsv(hs_macr_strain_table, celltype = "macrophage")

hs_macr_strain_sig <- extract_significant_genes(
  hs_macr_strain_table,
  excel = glue("excel/macrophage_de/macrophageonly_onlystrain_sig-v{ver}.xlsx"))
hs_macr_strain_sig
hs_macr_strain_highsig <- extract_significant_genes(
  hs_macr_strain_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/macrophageonly_onlystrain_highsig-v{ver}.xlsx"))
hs_macr_strain_highsig
```

#### Compare Drug and Strain Effects

Now let us consider the above two comparisons together.  First, I will
plot the logFC values of them against each other (drug on x-axis and
strain on the y-axis).  Then we can extract the significant genes in a
few combined categories of interest.  I assume these will focus
exclusively on the categories which include the introduction of the
drug.

```{r}
drug_strain_comp_df <- merge(hs_macr_drug_table[["data"]][["drug"]],
                             hs_macr_strain_table[["data"]][["strain"]],
                             by = "row.names")
drug_strain_comp_plot <- plot_linear_scatter(
  drug_strain_comp_df[, c("deseq_logfc.x", "deseq_logfc.y")])
## Contrasts: antimony/none, z23/z22; x-axis: drug, y-axis: strain
## top left: higher no drug, z23; top right: higher drug z23
## bottom left: higher no drug, z22; bottom right: higher drug z22
drug_strain_comp_plot$scatter
```

As I noted in the comments above, some quadrants of the scatter plot
are likely to be of greater interest to us than others (the right
side).  Because I get confused sometimes, the following block will
explicitly name the categories of likely interest, then ask which
genes are shared among them, and finally use UpSetR to extract the
various gene intersection/union categories.

```{r}
higher_drug <- hs_macr_drug_sig[["deseq"]][["downs"]][[1]]
higher_nodrug <- hs_macr_drug_sig[["deseq"]][["ups"]][[1]]
higher_z23 <- hs_macr_strain_sig[["deseq"]][["ups"]][[1]]
higher_z22 <- hs_macr_strain_sig[["deseq"]][["downs"]][[1]]
sum(rownames(higher_drug) %in% rownames(higher_z23))
sum(rownames(higher_drug) %in% rownames(higher_z22))
sum(rownames(higher_nodrug) %in% rownames(higher_z23))
sum(rownames(higher_nodrug) %in% rownames(higher_z22))

drug_z23_lst <- list("drug" = rownames(higher_drug),
                     "z23" = rownames(higher_z23))
upset_input <- UpSetR::fromList(drug_z23_lst)
higher_drug_z23 <- upset(upset_input, text.scale = 2)
higher_drug_z23

drug_z23_shared_genes <- overlap_groups(drug_z23_lst)
shared_genes_drug_z23 <- overlap_geneids(drug_z23_shared_genes, "drug:z23")
shared_genes_drug_z23 <- attr(drug_z23_shared_genes, "elements")[drug_z23_shared_genes[["drug:z23"]]]

drug_z22_lst <- list("drug" = rownames(higher_drug),
                     "z22" = rownames(higher_z22))
higher_drug_z22 <- upset(UpSetR::fromList(drug_z22_lst), text.scale = 2)
higher_drug_z22

drug_z22_shared_genes <- overlap_groups(drug_z22_lst)
shared_genes_drug_z22 <- overlap_geneids(drug_z22_shared_genes, "drug:z22")
shared_genes_drug_z22 <- attr(drug_z22_shared_genes, "elements")[drug_z22_shared_genes[["drug:z22"]]]
```

#### Perform gProfiler on drug/strain effect shared genes

Now that we have some populations of genes which are shared across the
drug/strain effects, let us pass them to some GSEA analyses and see
what pops out.

```{r}
wanted <- drug_z23_shared_genes[["drug:z23"]]
shared_genes_drug_z23 <- attr(drug_z23_shared_genes, "elements")[wanted]
shared_drug_z23_gp <- simple_gprofiler(shared_genes_drug_z23)
shared_drug_z23_gp[["pvalue_plots"]][["MF"]]
shared_drug_z23_gp[["pvalue_plots"]][["BP"]]
shared_drug_z23_gp[["pvalue_plots"]][["REAC"]]

wanted <- drug_z22_shared_genes[["drug:z22"]]
shared_genes_drug_z22 <- attr(drug_z22_shared_genes, "elements")[wanted]
shared_drug_z22_gp <- simple_gprofiler(shared_genes_drug_z22)
shared_drug_z22_gp[["pvalue_plots"]][["BP"]]
```

## Our main question of interest

The data structure hs_macr contains our primary macrophages, which
are, as shown above, the data we can really sink our teeth into.

Note, we expect some errors when running the combine_de_tables()
because not all methods I use are comfortable using the ratio or
ratios contrasts we added in the 'extras' argument.  As a result, when
we combine them into the larger output tables, those peculiar
contrasts fail.  This does not stop it from writing the rest of the
results, however.

```{r}
#test = deseq_pairwise(normalize_expt(hs_macr, filter=TRUE),
#                      model_batch = "svaseq", filter = TRUE,
#                      extra_contrasts = tmrc2_human_extra)

hs_macr_de <- all_pairwise(hs_macr, model_batch = "svaseq",
                           filter = TRUE, extra_contrasts = tmrc2_human_extra)
hs_macr_de
hs_single_table <- combine_de_tables(
  hs_macr_de, keepers = single_tmrc2_keeper,
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_z22sb_sb-v{ver}.xlsx"))
hs_single_table
hs_macr_table <- combine_de_tables(
  hs_macr_de, keepers = tmrc2_human_keepers,
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_table_macr_only-v{ver}.xlsx"))
hs_macr_table
combined_to_tsv(hs_macr_table, "macrophage")

hs_macr_sig <- extract_significant_genes(
  hs_macr_table,
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_sig-v{ver}.xlsx"))
hs_macr_sig
hs_macr_highsig <- extract_significant_genes(
  hs_macr_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_highsig-v{ver}.xlsx"))
hs_macr_highsig
```

## gene group upset

```{r}
nodrug_upset <- upsetr_combined_de(hs_macr_table,
                                   desired_contrasts = c("z22nosb_vs_uninf", "z23nosb_vs_uninf"))
pp(file = "images/nodrug_upset.png")
nodrug_upset
dev.off()
drug_upset <- upsetr_combined_de(hs_macr_table,
                                 desired_contrasts = c("z22sb_vs_sb", "z23sb_vs_sb"))
pp(file = "images/drug_upset.png")
drug_upset
dev.off()
```

# Significance barplot of interest

Olga kindly sent a set of particularly interesting contrasts and
colors for a significance barplot, they include the following:

* z2.3 vs. uninfected.
* z2.2 vs. uninfected.
* z2.3 vs z2.2
* z2.3Sbv vs z2.3
* z2.2Sbv vs z2.2
* z2.3Sbv vs z2.2Sbv
* Sbv vs uninfected.

The existing set of 'keepers' exvised to these is taken from the
extant set of 'tmrc2_human_keepers' and is as follows:


```{r}
barplot_keepers <- list(
  ## z2.3 vs uninfected
  "z23nosb_vs_uninf" = c("infz23", "uninfnone"),
  ## z2.2 vs uninfected
  "z22nosb_vs_uninf" = c("infz22", "uninfnone"),
  ## z2.3 vs z2.2
  "z23nosb_vs_z22nosb" = c("infz23", "infz22"),
  ## z2.3Sbv vs z2.3
  "z23sb_vs_z23nosb" = c("infsbz23", "infz23"),
  ## z2.2Sbv vs z2.2
  "z22sb_vs_z22nosb" = c("infsbz22", "infz22"),
  ## z2.3Sbv vs z2.2Sbv
  "z23sb_vs_z22sb" = c("infsbz23", "infsbz22"),
  ## Sbv vs uninfected.
  "sb_vs_uninf" = c("uninfsbnone", "uninfnone"))
barplot_combined <- combine_de_tables(
  hs_macr_de, keepers = barplot_keepers,
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_7contrasts-v{ver}.xlsx"))
```

Now let us use the colors suggested by Olga to make a barplot of
these...

```{r}
color_list <-  c( "#de8bf9", "#ad07e3","#410257", "#ffa0a0", "#f94040", "#a00000")
barplot_sig <- extract_significant_genes(
  barplot_combined, color_list = color_list, according_to = "deseq",
  excel = glue("excel/macrophage_de/hs_macr_drug_zymo_7contrasts_sig-v{ver}.xlsx"))
barplot_sig
```

# PROPER

In our last meeting there were some questions about the statistical
power of different future experimental designs.  One thing I can do is
to use PROPER to estimate the power of an extant dataset and infer
from that the likely power of other designs.

In order to use proper, one must feed it one or more DE tables.

```{r}
power_estimate <- simple_proper(hs_single_table)
power_estimate

power_estimate[[1]][["power_plot"]]
power_estimate[[1]][["powertd_plot"]]
power_estimate[[1]][["powerfd_plot"]]

```

### Our main questions in U937

Let us do the same comparisons in the U937 samples, though I will not
do the extra contrasts, primarily because I think the dataset is less
likely to support them.

```{r}
u937_de <- all_pairwise(u937_expt, model_batch = "svaseq",
                        filter = TRUE, do_noiseq = FALSE)
u937_de
u937_table <- combine_de_tables(
  u937_de, keepers = u937_keepers,
  excel = glue("excel/macrophage_de/u937_drug_zymo_table-v{ver}.xlsx"))
u937_table
combined_to_tsv(u937_table, celltype = "u937")

u937_sig <- extract_significant_genes(
  u937_table,
  excel = glue("excel/macrophage_de/u937_drug_zymo_sig-v{ver}.xlsx"))
u937_sig
u937_highsig <- extract_significant_genes(
  u937_table, min_mean_exprs = high_expression, exprs_column = high_expression_column,
  excel = glue("excel/macrophage_de/u937_drug_zymo_highsig-v{ver}.xlsx"))
u937_highsig
```

#### Compare (no)Sb z2.3/z2.2 treatments among macrophages

```{r}
upset_plots_hs_macr <- upsetr_sig(
  hs_macr_sig, both = TRUE,
  contrasts = c("z23sb_vs_z22sb", "z23nosb_vs_z22nosb"))
upset_plots_hs_macr[["both"]]
groups <- upset_plots_hs_macr[["both_groups"]]
shared_genes <- attr(groups, "elements")[groups[[2]]] %>%
  gsub(pattern = "^gene:", replacement = "")
length(shared_genes)

shared_gp <- simple_gprofiler(shared_genes)
shared_gp[["pvalue_plots"]][["MF"]]
shared_gp[["pvalue_plots"]][["BP"]]
shared_gp[["pvalue_plots"]][["REAC"]]

drug_genes <- attr(groups, "elements")[groups[["z23sb_vs_z22sb"]]] %>%
  gsub(pattern = "^gene:", replacement = "")
drugonly_gp <- simple_gprofiler(drug_genes)
drugonly_gp[["pvalue_plots"]][["BP"]]
```

I want to try something, directly include the u937 data in this...

```{r}
both_sig <- hs_macr_sig
names(both_sig[["deseq"]][["ups"]]) <- paste0("macr_", names(both_sig[["deseq"]][["ups"]]))
names(both_sig[["deseq"]][["downs"]]) <- paste0("macr_", names(both_sig[["deseq"]][["downs"]]))
u937_deseq <- u937_sig[["deseq"]]
names(u937_deseq[["ups"]]) <- paste0("u937_", names(u937_deseq[["ups"]]))
names(u937_deseq[["downs"]]) <- paste0("u937_", names(u937_deseq[["downs"]]))
both_sig[["deseq"]][["ups"]] <- c(both_sig[["deseq"]][["ups"]], u937_deseq[["ups"]])
both_sig[["deseq"]][["downs"]] <- c(both_sig[["deseq"]][["ups"]], u937_deseq[["downs"]])
summary(both_sig[["deseq"]][["ups"]])

upset_plots_both <- upsetr_sig(
  both_sig, both = TRUE,
  contrasts = c("macr_z23sb_vs_z22sb", "macr_z23nosb_vs_z22nosb",
                "u937_z23sb_vs_z22sb", "u937_z23nosb_vs_z22nosb"))
upset_plots_both$both
```

#### Compare DE results from macrophages and U937 samples

Looking a bit more closely at these, I think the u937 data is too
sparse to effectively compare.

```{r}
macr_u937_comparison <- compare_de_results(hs_macr_table, u937_table)
macr_u937_comparison$lfc_heat

macr_u937_venns <- compare_significant_contrasts(hs_macr_sig, second_sig_tables = u937_sig,
                                                 contrasts = "z23sb_vs_z23nosb")
macr_u937_venns$up_plot
macr_u937_venns$down_plot

macr_u937_venns_v2 <- compare_significant_contrasts(hs_macr_sig, second_sig_tables = u937_sig,
                                                    contrasts = "z22sb_vs_z22nosb")
macr_u937_venns_v2$up_plot
macr_u937_venns_v2$down_plot

macr_u937_venns_v3 <- compare_significant_contrasts(hs_macr_sig, second_sig_tables = u937_sig,
                                                    contrasts = "sb_vs_uninf")
macr_u937_venns_v3$up_plot
macr_u937_venns_v3$down_plot
```

### Compare macrophage/u937 with respect to z2.3/z2.2

```{r}
comparison_df <- merge(hs_macr_table[["data"]][["z23sb_vs_z22sb"]],
                       u937_table[["data"]][["z23sb_vs_z22sb"]],
                       by = "row.names")
macru937_z23z22_plot <- plot_linear_scatter(comparison_df[, c("deseq_logfc.x", "deseq_logfc.y")])
macru937_z23z22_plot$scatter

comparison_df <- merge(hs_macr_table[["data"]][["z23nosb_vs_z22nosb"]],
                       u937_table[["data"]][["z23nosb_vs_z22nosb"]],
                       by = "row.names")
macru937_z23z22_plot <- plot_linear_scatter(comparison_df[, c("deseq_logfc.x", "deseq_logfc.y")])
macru937_z23z22_plot$scatter
```

#### Add donor to the contrasts, no sva

```{r}
no_power_fact <- paste0(pData(hs_macr)[["donor"]], "_",
                        pData(hs_macr)[["condition"]])
table(pData(hs_macr)[["donor"]])
table(no_power_fact)
hs_nopower <- set_expt_conditions(hs_macr, fact = no_power_fact)
hs_nopower <- subset_expt(hs_nopower, subset="macrophagezymodeme!='none'")
hs_nopower_nosva_de <- all_pairwise(hs_nopower, model_batch = FALSE, filter = TRUE)
nopower_keepers <- list(
  "d01_zymo" = c("d01infz23", "d01infz22"),
  "d01_sbzymo" = c("d01infsbz23", "d01infsbz22"),
  "d02_zymo" = c("d02infz23", "d02infz22"),
  "d02_sbzymo" = c("d02infsbz23", "d02infsbz22"),
  "d09_zymo" = c("d09infz23", "d09infz22"),
  "d09_sbzymo" = c("d09infsbz23", "d09infsbz22"),
  "d81_zymo" = c("d81infz23", "d81infz22"),
  "d81_sbzymo" = c("d81infsbz23", "d81infsbz22"))
hs_nopower_nosva_table <- combine_de_tables(
  hs_nopower_nosva_de, keepers = nopower_keepers,
  excel = glue("excel/macrophage_de/hs_nopower_table-v{ver}.xlsx"))
##                                  extra_contrasts = extra)
hs_nopower_nosva_sig <- extract_significant_genes(
  hs_nopower_nosva_table,
  excel = glue("excel/macrophage_de/hs_nopower_nosva_sig-v{ver}.xlsx"))

d01d02_zymo_nosva_comp <- merge(hs_nopower_nosva_table[["data"]][["d01_zymo"]],
                                hs_nopower_nosva_table[["data"]][["d02_zymo"]],
                                by="row.names")
d0102_zymo_nosva_plot <- plot_linear_scatter(d01d02_zymo_nosva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0102_zymo_nosva_plot$scatter
d0102_zymo_nosva_plot$correlation
d0102_zymo_nosva_plot$lm_rsq

d09d81_zymo_nosva_comp <- merge(hs_nopower_nosva_table[["data"]][["d09_zymo"]],
                                hs_nopower_nosva_table[["data"]][["d81_zymo"]],
                                by="row.names")
d0981_zymo_nosva_plot <- plot_linear_scatter(d09d81_zymo_nosva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0981_zymo_nosva_plot$scatter
d0981_zymo_nosva_plot$correlation
d0981_zymo_nosva_plot$lm_rsq

d01d81_zymo_nosva_comp <- merge(hs_nopower_nosva_table[["data"]][["d01_zymo"]],
                                hs_nopower_nosva_table[["data"]][["d81_zymo"]],
                                by="row.names")
d0181_zymo_nosva_plot <- plot_linear_scatter(d01d81_zymo_nosva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0181_zymo_nosva_plot$scatter
d0181_zymo_nosva_plot$correlation
d0181_zymo_nosva_plot$lm_rsq

upset_plots_nosva <- upsetr_sig(hs_nopower_nosva_sig, both=TRUE,
                                contrasts=c("d01_zymo", "d02_zymo", "d09_zymo", "d81_zymo"))
upset_plots_nosva$up
upset_plots_nosva$down
upset_plots_nosva$both
## The 7th element in the both groups list is the set shared among all donors.
## I don't feel like writing out x:y:z:a
groups <- upset_plots_nosva[["both_groups"]]
shared_genes <- attr(groups, "elements")[groups[[7]]] %>%
  gsub(pattern = "^gene:", replacement = "")
shared_gp <- simple_gprofiler(shared_genes)
shared_gp$pvalue_plots$MF
shared_gp$pvalue_plots$BP
shared_gp$pvalue_plots$REAC
shared_gp$pvalue_plots$WP
```

#### Add donor to the contrasts, sva

```{r}
hs_nopower_sva_de <- all_pairwise(hs_nopower, model_batch = "svaseq", filter = TRUE)
nopower_keepers <- list(
  "d01_zymo" = c("d01infz23", "d01infz22"),
  "d01_sbzymo" = c("d01infsbz23", "d01infsbz22"),
  "d02_zymo" = c("d02infz23", "d02infz22"),
  "d02_sbzymo" = c("d02infsbz23", "d02infsbz22"),
  "d09_zymo" = c("d09infz23", "d09infz22"),
  "d09_sbzymo" = c("d09infsbz23", "d09infsbz22"),
  "d81_zymo" = c("d81infz23", "d81infz22"),
  "d81_sbzymo" = c("d81infsbz23", "d81infsbz22"))
hs_nopower_sva_table <- combine_de_tables(
  hs_nopower_sva_de, keepers = nopower_keepers,
  excel = glue("excel/macrophage_de/hs_nopower_table-v{ver}.xlsx"))
##                                  extra_contrasts = extra)
hs_nopower_sva_sig <- extract_significant_genes(
  hs_nopower_sva_table,
  excel = glue("excel/macrophage_de/hs_nopower_sva_sig-v{ver}.xlsx"))

d01d02_zymo_sva_comp <- merge(hs_nopower_sva_table[["data"]][["d01_zymo"]],
                              hs_nopower_sva_table[["data"]][["d02_zymo"]],
                              by="row.names")
d0102_zymo_sva_plot <- plot_linear_scatter(d01d02_zymo_sva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0102_zymo_sva_plot$scatter
d0102_zymo_sva_plot$correlation
d0102_zymo_sva_plot$lm_rsq

d09d81_zymo_sva_comp <- merge(hs_nopower_sva_table[["data"]][["d09_zymo"]],
                              hs_nopower_sva_table[["data"]][["d81_zymo"]],
                              by="row.names")
d0981_zymo_sva_plot <- plot_linear_scatter(d09d81_zymo_sva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0981_zymo_sva_plot$scatter
d0981_zymo_sva_plot$correlation
d0981_zymo_sva_plot$lm_rsq

d01d81_zymo_sva_comp <- merge(hs_nopower_sva_table[["data"]][["d01_zymo"]],
                              hs_nopower_sva_table[["data"]][["d81_zymo"]],
                              by="row.names")
d0181_zymo_sva_plot <- plot_linear_scatter(d01d81_zymo_sva_comp[, c("deseq_logfc.x", "deseq_logfc.y")])
d0181_zymo_sva_plot$scatter
d0181_zymo_sva_plot$correlation
d0181_zymo_sva_plot$lm_rsq

upset_plots_sva <- upsetr_sig(hs_nopower_sva_sig, both=TRUE,
                              contrasts=c("d01_zymo", "d02_zymo", "d09_zymo", "d81_zymo"))
upset_plots_sva$up
upset_plots_sva$down
upset_plots_sva$both
## The 7th element in the both groups list is the set shared among all donors.
## I don't feel like writing out x:y:z:a
groups <- upset_plots_sva[["both_groups"]]
shared_genes <- attr(groups, "elements")[groups[[7]]] %>%
  gsub(pattern = "^gene:", replacement = "")
shared_gp <- simple_gprofiler(shared_genes)
shared_gp$pvalue_plots$MF
shared_gp$pvalue_plots$BP
shared_gp$pvalue_plots$REAC
shared_gp$pvalue_plots$WP
```

### Donor comparison

```{r}
hs_donors <- set_expt_conditions(hs_macr, fact = "donor")
donor_de <- all_pairwise(hs_donors, model_batch="svaseq", filter=TRUE)
donor_de
donor_table <- combine_de_tables(
  donor_de,
  excel=glue("excel/macrophage_de/donor_tables-v{ver}.xlsx"))
donor_table
donor_sig <- extract_significant_genes(
  donor_table,
  excel = glue("excel/macrophage_de/donor_sig-v{ver}.xlsx"))
donor_sig
```

#### Primary query contrasts

The final contrast in this list is interesting because it depends on
the extra contrasts applied to the all_pairwise() above.  In my way of
thinking, the primary comparisons to consider are either cross-drug or
cross-strain, but not both.  However I think in at least a few
instances Olga is interested in strain+drug / uninfected+nodrug.

#### Write contrast results

Now let us write out the xlsx file containing the above contrasts.
The file with the suffix _table-version will therefore contain all
genes and the file with the suffix _sig-version will contain only
those deemed significant via our default criteria of DESeq2 |logFC| >= 1.0
and adjusted p-value <= 0.05.

# Over representation searches

I decided to make one initially small, but I think quickly big change
to the organization of this document:  I am moving the GSEA searches
up to immediately after the DE.  I will then move the plots of the
gprofiler results to immediately after the various volcano plots so
that it is easier to interpret them.

```{r}
all_gp <- all_gprofiler(hs_macr_sig)
for (g in seq_len(length(all_gp))) {
  name <- names(all_gp)[g]
  datum <- all_gp[[name]]
  filename <- glue("excel/macrophage_de/gprofiler/{name}_gprofiler-v{ver}.xlsx")
  written <- sm(write_gprofiler_data(datum, excel = filename))
}
```

# Plot contrasts of interest

One suggestion I received recently was to set the axes for these
volcano plots to be static rather than let ggplot choose its own.  I
am assuming this is only relevant for pairs of contrasts, but that
might not be true.

## Individual zymodemes vs. uninfected

### Infected with z2.3 no Antimonial vs. Uninfected

```{r}
plot_colors <- get_expt_colors(hs_macr_table[["input"]][["input"]])
x_limits <- c(-20, 10)

## The original plot from my xlsx file
hs_macr_table$plots$z23nosb_vs_uninf$deseq_vol_plots

z23nosb_vs_uninf_volcano <- plot_volcano_condition_de(
  input = hs_macr_table[["data"]][["z23nosb_vs_uninf"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["uninfnone"]], color_low = plot_colors[["infz23"]])
z23nosb_vs_uninf_volcano$plot +
  scale_x_continuous(limits = x_limits)
plotly::ggplotly(z23nosb_vs_uninf_volcano$plot)

z23nosb_vs_uninf_volcano_nol <- plot_volcano_condition_de(
  input = hs_macr_table[["data"]][["z23nosb_vs_uninf"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = NULL, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["uninfnone"]], color_low = plot_colors[["infz23"]])
z23nosb_vs_uninf_volcano_nol$plot +
  scale_x_continuous(limits = x_limits)

all_gp[["z23nosb_vs_uninf_up"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_uninf_up"]][["pvalue_plots"]][["KEGG"]]
## KEGG, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_uninf_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_uninf_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_uninf_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_uninf_up"]][["interactive_plots"]][["WP"]]

all_gp[["z23nosb_vs_uninf_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23nosb_vs_uninf_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23nosb_vs_uninf_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```

### Infected with z2.2 no Antimonial vs. Uninfected

```{r}
## The original plot
hs_macr_table$plots$z22nosb_vs_uninf$deseq_vol_plots

z22nosb_vs_uninf_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z22nosb_vs_uninf"]], "z22nosb_vs_uninf",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["uninfnone"]], color_low = plot_colors[["infz22"]])
z22nosb_vs_uninf_volcano$plot +
  scale_x_continuous(limits = x_limits)
plotly::ggplotly(z22nosb_vs_uninf_volcano$plot)

z22nosb_vs_uninf_volcano_nol <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z22nosb_vs_uninf"]], "z22nosb_vs_uninf",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = NULL, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["uninfnone"]], color_low = plot_colors[["infz22"]])
z22nosb_vs_uninf_volcano_nol$plot +
  scale_x_continuous(limits = x_limits)

all_gp[["z22nosb_vs_uninf_up"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.2 without drug vs. uninfected without drug, up.
all_gp[["z22nosb_vs_uninf_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.2 without drug vs. uninfected without drug, up.
all_gp[["z22nosb_vs_uninf_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.2 without drug vs. uninfected without drug, up.
all_gp[["z22nosb_vs_uninf_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.2 without drug vs. uninfected without drug, up.

all_gp[["z22nosb_vs_uninf_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.2 without drug vs. uninfected without drug, down.
all_gp[["z22nosb_vs_uninf_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.2 without drug vs. uninfected without drug, down.
all_gp[["z22nosb_vs_uninf_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```

### Infected with z2.3 treated vs. Uninfected treated

```{r}
## The original plot
hs_macr_table$plots$z23sb_vs_sb$deseq_vol_plots

z23sb_vs_uninfsb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_sb"]], "z23sb_vs_sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["infsbz23"]], color_low = plot_colors[["uninfsbnone"]])
z23sb_vs_uninfsb_volcano$plot +
  scale_x_continuous(limits = x_limits)
plotly::ggplotly(z23sb_vs_uninfsb_volcano$plot)

z23sb_vs_uninfsb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_sb"]], "z23sb_vs_sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = NULL, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["infsbz23"]], color_low = plot_colors[["uninfsbnone"]])
z23sb_vs_uninfsb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.3 untreated vs. z2.2 untreated

```{r}
## The original plot
hs_macr_table$plots$z23nosb_vs_z22nosb$deseq_vol_plots

z23nosb_vs_z22nosb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23nosb_vs_z22nosb"]], "z23nosb_vs_z22nosb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_high = plot_colors[["infz23"]], color_low = plot_colors[["infz22"]])
z23nosb_vs_z22nosb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.3 treated vs. z2.2 treated

```{r}
## The original plot
hs_macr_table$plots$z23sb_vs_z22sb$deseq_vol_plots

z23sb_vs_z22sb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_z22sb"]], "z23sb_vs_z22sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = FALSE,
  color_high = plot_colors[["infsbz23"]], color_low = plot_colors[["infsbz22"]])
z23sb_vs_z22sb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.3 SB treated vs. z2.3 untreated

```{r}
## The original plot
hs_macr_table$plots$z23sb_vs_z23nosb$deseq_vol_plots

z23sb_vs_z23nosb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_z23nosb"]], "z23sb_vs_z23nosb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_low = plot_colors[["infsbz23"]], color_high = plot_colors[["infz23"]])
z23sb_vs_z23nosb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.3 SB treated vs. z2.3 untreated

```{r}
## The original plot
hs_macr_table$plots$z22sb_vs_z22nosb$deseq_vol_plots

z22sb_vs_z22nosb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z22sb_vs_z22nosb"]], "z22sb_vs_z22nosb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_low = plot_colors[["infsbz22"]], color_high = plot_colors[["infz22"]])
z22sb_vs_z22nosb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.3 SB treated vs. uninfected treated

```{r}
## The original plot
hs_macr_table$plots$z23sb_vs_sb$deseq_vol_plots

z23sb_vs_sb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_sb"]], "z23sb_vs_sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_low = plot_colors[["infsbz23"]], color_high = plot_colors[["uninfsbnone"]])
z23sb_vs_sb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

### Infected with z2.2 SB treated vs. uninfected treated

```{r}
## The original plot
hs_macr_table$plots$z22sb_vs_sb$deseq_vol_plots

z22sb_vs_sb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z22sb_vs_sb"]], "z22sb_vs_sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol", invert = TRUE,
  color_low = plot_colors[["infsbz22"]], color_high = plot_colors[["uninfsbnone"]])
z22sb_vs_sb_volcano$plot +
  scale_x_continuous(limits = x_limits)
```

Check that my perception of the number of significant up/down genes
matches what the table/venn says.

```{r}
shared <- Vennerable::Venn(list("drug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z23sb_vs_uninf"]]),
                                "nodrug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z23nosb_vs_uninf"]])))
pp(file="images/z23_vs_uninf_venn_up.png")
Vennerable::plot(shared)
dev.off()
Vennerable::plot(shared)
## I see 910 z23sb/uninf and 670 no z23nosb/uninf genes in the venn diagram.
length(shared@IntersectionSets[["10"]]) + length(shared@IntersectionSets[["11"]])
dim(hs_macr_sig[["deseq"]][["ups"]][["z23sb_vs_uninf"]])

shared <- Vennerable::Venn(list("drug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z22sb_vs_uninf"]]),
                                "nodrug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z22nosb_vs_uninf"]])))
pp(file="images/z22_vs_uninf_venn_up.png")
Vennerable::plot(shared)
dev.off()
Vennerable::plot(shared)

length(shared@IntersectionSets[["10"]]) + length(shared@IntersectionSets[["11"]])
dim(hs_macr_sig[["deseq"]][["ups"]][["z22sb_vs_uninf"]])
```

*Note to self*: There is an error in my volcano plot code which takes
effect when the numerator and denominator of the all_pairwise
contrasts are different than those in combine_de_tables.  It is
putting the ups/downs on the correct sides of the plot, but calling
the down genes 'up' and vice-versa.  The reason for this is that I did
a check for this happening, but used the wrong argument to handle it.

A likely bit of text for these volcano plots:

The set of genes differentially expressed between the zymodeme 2.3
and uninfected samples without druge treatment was quantified with
DESeq2 and included surrogate estimates from SVA.  Given the criteria
of significance of a abs(logFC) >= 1.0 and false discovery rate
adjusted p-value <= 0.05, 670 genes were observed as significantly
increased between the infected and uninfected samples and 386 were
observed as decreased. The most increased genes from the uninfected
samples include some which are potentially indicative of a strong
innate immune response and the inflammatory response.

In contrast, when the set of genes differentially expressed between
the zymodeme 2.2 and uninfected samples was visualized, only 7 genes
were observed as decreased and 435 increased.  The inflammatory
response was significantly less apparent in this set, but instead
included genes related to transporter activity and oxidoreductases.

## Direct zymodeme comparisons

An orthogonal comparison to that performed above is to directly
compare the zymodeme 2.3 and 2.2 samples with and without antimonial
treatment.

```{r}
z23nosb_vs_z22nosb_volcano <- plot_volcano_de(
  table = hs_macr_table[["data"]][["z23nosb_vs_z22nosb"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  shapes_by_state = FALSE, color_by = "fc",  label = 10, label_column = "hgncsymbol")
plotly::ggplotly(z23nosb_vs_z22nosb_volcano$plot)

z23sb_vs_z22sb_volcano <- plot_volcano_de(
  table = hs_macr_table[["data"]][["z23sb_vs_z22sb"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  shapes_by_state = FALSE, color_by = "fc",  label = 10, label_column = "hgncsymbol")
plotly::ggplotly(z23sb_vs_z22sb_volcano$plot)
```

```{r}
z23nosb_vs_z22nosb_volcano$plot +
  xlim(-10, 10) +
  ylim(0, 60)

pp(file="images/z23nosb_vs_z22nosb_reactome_up.png", image=all_gp[["z23nosb_vs_z22nosb_up"]][["pvalue_plots"]][["REAC"]], height=12, width=9)
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_z22nosb_up"]][["pvalue_plots"]][["KEGG"]]
## KEGG, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_z22nosb_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_z22nosb_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_z22nosb_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23nosb_vs_z22nosb_up"]][["interactive_plots"]][["WP"]]

all_gp[["z23nosb_vs_z22nosb_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23nosb_vs_z22nosb_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23nosb_vs_z22nosb_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```

```{r}
z23sb_vs_z22sb_volcano$plot +
  xlim(-10, 10) +
  ylim(0, 60)

pp(file="images/z23sb_vs_z22sb_reactome_up.png", image=all_gp[["z23sb_vs_z22sb_up"]][["pvalue_plots"]][["REAC"]], height=12, width=9)
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z22sb_up"]][["pvalue_plots"]][["KEGG"]]
## KEGG, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z22sb_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z22sb_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z22sb_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z22sb_up"]][["interactive_plots"]][["WP"]]

all_gp[["z23sb_vs_z22sb_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23sb_vs_z22sb_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23sb_vs_z22sb_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```


```{r}
shared <- Vennerable::Venn(list("drug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z23sb_vs_z22sb"]]),
                                "nodrug" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z23nosb_vs_z22nosb"]])))
pp(file="images/drug_nodrug_venn_up.png")
Vennerable::plot(shared)
dev.off()
Vennerable::plot(shared)

shared <- Vennerable::Venn(list("drug" = rownames(hs_macr_sig[["deseq"]][["downs"]][["z23sb_vs_z22sb"]]),
                                "nodrug" = rownames(hs_macr_sig[["deseq"]][["downs"]][["z23nosb_vs_z22nosb"]])))
pp(file="images/drug_nodrug_venn_down.png")
Vennerable::plot(shared)
dev.off()
```

A slightly different way of looking at the differences between the two
zymodeme infections is to directly compare the infected samples with
and without drug.  Thus, when a volcano plot showing the comparison of
the zymodeme 2.3 vs. 2.2 samples was plotted, 484 genes were observed
as increased and 422 decreased; these groups include many of the same
inflammatory (up) and membrane (down) genes.

Similar patterns were observed when the antimonial was included.
Thus, when a Venn diagram of the two sets of increased genes was
plotted, a significant number of the genes was observed as increased
(313) and decreased (244) in both the untreated and antimonial treated
samples.

## Drug effects on each zymodeme infection

Another likely question is to directly compare the treated vs
untreated samples for each zymodeme infection in order to visualize
the effects of antimonial.

```{r}
z23sb_vs_z23nosb_volcano <- plot_volcano_de(
  table = hs_macr_table[["data"]][["z23sb_vs_z23nosb"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  shapes_by_state = FALSE, color_by = "fc",  label = 10, label_column = "hgncsymbol")
plotly::ggplotly(z23sb_vs_z23nosb_volcano$plot)
z22sb_vs_z22nosb_volcano <- plot_volcano_de(
  table = hs_macr_table[["data"]][["z22sb_vs_z22nosb"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  shapes_by_state = FALSE, color_by = "fc",  label = 10, label_column = "hgncsymbol")
plotly::ggplotly(z22sb_vs_z22nosb_volcano$plot)
```

```{r}
z23sb_vs_z23nosb_volcano$plot +
  xlim(-8, 8) +
  ylim(0, 210)

pp(file="images/z23sb_vs_z23nosb_reactome_up.png",
   image=all_gp[["z23sb_vs_z23nosb_up"]][["pvalue_plots"]][["REAC"]], height=12, width=9)
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z23nosb_up"]][["pvalue_plots"]][["KEGG"]]
## KEGG, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z23nosb_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z23nosb_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z23nosb_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z23sb_vs_z23nosb_up"]][["interactive_plots"]][["WP"]]

all_gp[["z23sb_vs_z23nosb_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23sb_vs_z23nosb_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z23sb_vs_z23nosb_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```

```{r}
z22sb_vs_z22nosb_volcano$plot +
  xlim(-8, 8) +
  ylim(0, 210)

pp(file="images/z22sb_vs_z22nosb_reactome_up.png",
   image=all_gp[["z22sb_vs_z22nosb_up"]][["pvalue_plots"]][["REAC"]], height=12, width=9)
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z22sb_vs_z22nosb_up"]][["pvalue_plots"]][["KEGG"]]
## KEGG, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z22sb_vs_z22nosb_up"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z22sb_vs_z22nosb_up"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z22sb_vs_z22nosb_up"]][["pvalue_plots"]][["WP"]]
## WikiPathways, zymodeme2.3 without drug vs. uninfected without drug, up.
all_gp[["z22sb_vs_z22nosb_up"]][["interactive_plots"]][["WP"]]

all_gp[["z22sb_vs_z22nosb_down"]][["pvalue_plots"]][["REAC"]]
## Reactome, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z22sb_vs_z22nosb_down"]][["pvalue_plots"]][["MF"]]
## MF, zymodeme2.3 without drug vs. uninfected without drug, down.
all_gp[["z22sb_vs_z22nosb_down"]][["pvalue_plots"]][["TF"]]
## TF, zymodeme2.3 without drug vs. uninfected without drug, down.
```

```{r}
shared <- Vennerable::Venn(list("z23" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z23sb_vs_z23nosb"]]),
                                "z22" = rownames(hs_macr_sig[["deseq"]][["ups"]][["z22sb_vs_z22nosb"]])))
pp(file="images/z23_z22_drug_venn_up.png")
Vennerable::plot(shared)
dev.off()
Vennerable::plot(shared)

shared <- Vennerable::Venn(list("z23" = rownames(hs_macr_sig[["deseq"]][["downs"]][["z23sb_vs_z23nosb"]]),
                                "z22" = rownames(hs_macr_sig[["deseq"]][["downs"]][["z22sb_vs_z22nosb"]])))
pp(file="images/z23_z22_drug_venn_down.png")
Vennerable::plot(shared)
dev.off()
Vennerable::plot(shared)
```

Note: I am settig the x and y-axis boundaries by allowing the plotter
to pick its own axis the first time, writing down the ranges I
observe, and then setting them to the largest of the pair.  It is
therefore possible that I missed one or more genes which lies outside
that range.

The previous plotted contrasts sought to show changes between the two
strains z2.3 and z2.2.  Conversely, the previous volcano plots seek to
directly compare each strain before/after drug treatment.

## LRT of the Human Macrophage

```{r}
tmrc2_lrt_strain_drug <- deseq_lrt(hs_macr, interactor_column = "drug",
                                   interest_column = "macrophagezymodeme", factors = c("drug", "macrophagezymodeme"))
tmrc2_lrt_strain_drug$cluster_data$plot
```

## Parasite

```{r}
lp_macrophage_de <- all_pairwise(lp_macrophage_nosb,
                                 model_batch="svaseq", filter=TRUE)
tmrc2_parasite_keepers <- list(
  "z23_vs_z22" = c("z23", "z22"))
lp_macrophage_table <- combine_de_tables(
  lp_macrophage_de, keepers = tmrc2_parasite_keepers,
  excel = glue("excel/macrophage_de/macrophage_parasite_infection_de-v{ver}.xlsx"))
lp_macrophage_sig <- extract_significant_genes(
  lp_macrophage_table,
  excel = glue("excel/macrophage_de/macrophage_parasite_sig-v{ver}.xlsx"))

lp_macrophage_table[["plots"]][["z23nosb_vs_z22nosb"]][["deseq_vol_plots"]][["plot"]]

up_genes <- lp_macrophage_sig[["deseq"]][["ups"]][[1]]
dim(up_genes)
down_genes <- lp_macrophage_sig[["deseq"]][["downs"]][[1]]
dim(down_genes)
```

```{r parasite_volcano}
lp_z23sb_vs_z22sb_volcano <- plot_volcano_de(
  table = lp_macrophage_table[["data"]][["z23_vs_z22"]],
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  shapes_by_state = FALSE, color_by = "fc",  label = 10, label_column = "hgncsymbol")
plotly::ggplotly(lp_z23sb_vs_z22sb_volcano$plot)
lp_z23sb_vs_z22sb_volcano$plot
```

```{r}
lp_lengths <- all_lp_annot[, c("gid", "annot_cds_length")]
colnames(lp_lengths)  <- c("ID", "length")

up_goseq <- simple_goseq(up_genes, go_db = lp_go, length_db = lp_lengths)
## View categories over represented in the 2.3 samples
up_goseq$pvalue_plots$bpp_plot_over
down_goseq <- simple_goseq(down_genes, go_db = lp_go, length_db = lp_lengths)
## View categories over represented in the 2.2 samples
down_goseq$pvalue_plots$bpp_plot_over

written_goseq <- write_goseq_data(up_goseq,
                                  excel = glue("lp_macrophage_increased_z2.3_goseq-v{ver}.xlsx"))
written_goseq <- write_goseq_data(down_goseq,
                                  excel = glue("lp_macrophage_increased_z2.2_goseq-v{ver}.xlsx"))
```

# GSVA

```{r}
hs_infected <- subset_expt(hs_macrophage, subset="macrophagetreatment!='uninf'") %>%
  subset_expt(subset="macrophagetreatment!='uninf_sb'")
hs_gsva_c2 <- simple_gsva(hs_infected)
hs_gsva_c2_meta <- get_msigdb_metadata(hs_gsva_c2, msig_xml="reference/msigdb_v7.2.xml")
hs_gsva_c2_sig <- get_sig_gsva_categories(hs_gsva_c2_meta, excel = "excel/macrophage_de/hs_macrophage_gsva_c2_sig.xlsx")
hs_gsva_c2_sig$raw_plot

hs_gsva_c7 <- simple_gsva(hs_infected, signature_category = "c7")
hs_gsva_c7_meta <- get_msigdb_metadata(hs_gsva_c7, msig_xml="reference/msigdb_v7.2.xml")
hs_gsva_c7_sig <- get_sig_gsva_categories(hs_gsva_c7, excel = "excel/macrophage_de/hs_macrophage_gsva_c7_sig.xlsx")
hs_gsva_c7_sig$raw_plot
```

# Try out a new tool

Two reasons: Najib loves him some PCA, this uses wikipathways, which is something I think is neat.

Ok, I spent some time looking through the code and I have some
problems with some of the design decisions.

Most importantly, it requires a data.frame() which has the following format:

1.  No rownames, instead column #1 is the sample ID.
2.  Columns 2-m are the categorical/survival/etc metrics.
3.  Columns m-n are 1 gene-per-column with log2 values.

But when I think about it I think I get the idea, they want to be able to do modelling stuff
more easily with response factors.

```{r}
library(pathwayPCA)
library(rWikiPathways)

downloaded <- downloadPathwayArchive(organism = "Homo sapiens", format = "gmt")
data_path <- system.file("extdata", package = "pathwayPCA")
wikipathways <- read_gmt(paste0(data_path, "/wikipathways_human_symbol.gmt"),
                         description = TRUE)

expt <- subset_expt(hs_macrophage, subset = "macrophagetreatment!='uninf'") %>%
  subset_expt(subset = "macrophagetreatment!='uninf_sb'")
expt <- set_expt_conditions(expt, fact = "macrophagezymodeme")

symbol_vector <- fData(expt)[[symbol_column]]
names(symbol_vector) <- rownames(fData(expt))
symbol_df <- as.data.frame(symbol_vector)

assay_df <- merge(symbol_df, as.data.frame(exprs(expt)), by = "row.names")
assay_df[["Row.names"]] <- NULL
rownames(assay_df) <- make.names(assay_df[["symbol_vector"]], unique = TRUE)
assay_df[["symbol_vector"]] <- NULL
assay_df <- as.data.frame(t(assay_df))
assay_df[["SampleID"]] <- rownames(assay_df)
assay_df <- dplyr::select(assay_df, "SampleID", everything())

factor_df <- as.data.frame(pData(expt))
factor_df[["SampleID"]] <- rownames(factor_df)
factor_df <- dplyr::select(factor_df, "SampleID", everything())
factor_df <- factor_df[, c("SampleID", factors)]

tt <- CreateOmics(
  assayData_df = assay_df,
  pathwayCollection_ls = wikipathways,
  response = factor_df,
  respType = "categorical",
  minPathSize=5)

super <- AESPCA_pVals(
  object = tt,
  numPCs = 2,
  parallel = FALSE,
  numCores = 8,
  numReps = 2,
  adjustment = "BH")
```

```{r}
## Stopping this because it takes forever
##if (!isTRUE(get0("skip_load"))) {
##  pander::pander(sessionInfo())
##  message("This is hpgltools commit: ", get_git_commit())
##  message("Saving to ", savefile)
##  tmp <- sm(saveme(filename = savefile))
##}
```

```{r loadme_after, eval = FALSE}
tmp <- loadme(filename = savefile)
```
