---
title: "Macrophage response to different strains: `r Sys.getenv('VERSION')`: README."
author: "atb abelew@gmail.com"
bibliography: /home/trey/scratch/zotero_library/atb.bib
date: "`r Sys.Date()`"
output:
  html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    number_sections: true
    self_contained: true
    theme: readable
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: false
  rmdformats::readthedown:
    code_download: true
    code_folding: show
    df_print: paged
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    width: 300
    keep_md: false
    mode: selfcontained
    toc_float: true
  BiocStyle::html_document:
    code_download: true
    code_folding: show
    fig_caption: true
    fig_height: 7
    fig_width: 7
    highlight: zenburn
    keep_md: false
    mode: selfcontained
    toc_float: true
---

<style type="text/css">
body, td {
  font-size: 16px;
}
code.r {
  font-size: 16px;
}
pre {
  font-size: 16px
}
body .main-container {
  max-width: 1600px;
}
</style>

```{r options, include=FALSE}
knitr::opts_knit$set(
  progress = TRUE, verbose = TRUE, width = 90, echo = TRUE)
knitr::opts_chunk$set(
  error = TRUE, fig.width = 8, fig.height = 8, fig.retina = 2,
  out.width = "100%", dev = "png",
  dev.args = list(png = list(type = "cairo-png")))
old_options <- options(digits = 4, stringsAsFactors = FALSE, knitr.duplicate.label = "allow")
ggplot2::theme_set(ggplot2::theme_bw(base_size = 12))
ver <- Sys.getenv("VERSION")
```

# Introduction

This document is intended to provide a roadmap to the figures which
were created by the various R markdown notebooks for this project and
which were included in "Interplay of human macrophage response and
natural resistance of L. (V.) panamensis to pentavalent antimony."

# Roadmap

## Figure 1B

This is a PCA plot colored by infection state, the source of this is
in 02visualization.Rmd, section 2.3.1 and may be found by searching
for the code:

```{r, eval=FALSE}
hs_macr_nb_pca <- plot_pca(hs_macr_nb)
pp(file = "figures/figure_1b_pca.svg")
hs_macr_nb_pca[["plot"]]
dev.off()
```

![PCA showing macrophage transcriptional responses](figures/figure_1b_pca.svg)

## Figure S1A

Variance Partition with donor+zymodeme+drug

This may be found in 02visualization in section 3 and is created in
the following stanza:

```{r, eval=FALSE}
donor_drug_zymo_varpart <- simple_varpart(
  hs_macr,
  factors = c("donor", "drug", "macrophagezymodeme"))
pp(file = "figures/donor_drug_zymodeme_varpart.svg")
donor_drug_zymo_varpart
dev.off()
```

![Variance Partition with donor, zymodeme, and drug](figures/donor_drug_zymodeme_varpart.svg)

## Figure S1B

PCA using donor, found in 02visualization.Rmd in the block containing:

```{r, eval=FALSE}
drug_donor_test <- set_expt_conditions(hs_macr, fact = "drug") %>%
  set_expt_batches(fact = "donor")
drug_donor_norm <- normalize_expt(drug_donor_test, norm = "quant", filter = TRUE,
                                  convert = "cpm", transform = "log2")
donor_pca <- plot_pca(drug_donor_norm)
```

![PCA highlighting donor variance](figures/donor_norm_pca.svg)

## Figure 2A: Volcano of z2.3 vs uninfected

A volcanot plot of uninfected vs z2.3 without drug treatment.  This
may be found in the following block:

```{r, eval=FALSE}
labeled <- z23nosb_vs_uninf_volcano[["plot"]] +
  scale_x_continuous(limits = c(-6, 22), breaks = c(-6, -4, -2, 0, 2, 4, 6, 8, 10, 21, 22)) +
  ggbreak::scale_x_break(c(10, 20), scales = 0.2, space = 0.02)
pp(file = "figures/fig2a_labeled_with_break.pdf")
labeled
dev.off()
labeled
```

![Volcano of no-drug z2.3 vs. uninfected](figures/fig2a_labeled_with_break.svg)

## Figure 2B: Volcano of z2.2 vs uninfected

A volcanot plot of uninfected vs z2.3 without drug treatment.  This
may be found in the following block:

```{r, eval=FALSE}
labeled <- z22nosb_vs_uninf_volcano[["plot"]] +
  scale_x_continuous(limits = c(-2, 21), breaks = c(-2, 0, 2, 4, 6, 8, 10, 21, 22)) +
  ggbreak::scale_x_break(c(11, 20), scales = 0.2, space = 0.02)
pp(file = "figures/fig2b_labeled_with_break.svg")
labeled
dev.off()
```

![Volcano of no-drug z2.2 vs. uninfected](figures/fig2b_labeled_with_break.svg)

## Figure 2C: Volcano of orange vs forest

A volcano comparing z2.3 and z2.2 without drug treatment:

```{r, eval=FALSE}
z23nosb_vs_z22nosb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23nosb_vs_z22nosb"]], "z23nosb_vs_z22nosb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol",
  color_low = plot_colors[["infz23"]], color_high = plot_colors[["infz22"]])

labeled <- z23nosb_vs_z22nosb_volcano[["plot"]] +
  scale_x_continuous(breaks = c(-10, -8, -6, -4, -2, 0, 2, 4, 6))

pp(file = "figures/fig2c_labeled.svg")
labeled
dev.off()
```

![Volcano of no-drug z2.3 vs. z2.2](figures/fig2c_labeled.svg)

## Figure 2D: UpSet plot showing unique and shared genes

The following upset plot seeks to show the groups of genes which are
unique and shared among the z2.3 and z2.2 comparisons to uninfected.

```{r, eval=FALSE}
nodrug_upset <- upsetr_combined_de(hs_macr_table,
                                   desired_contrasts = c("z22nosb_vs_uninf", "z23nosb_vs_uninf"))
pp(file = "figures/nodrug_upset.svg")
nodrug_upset
dev.off()
```

![Upset comparing z2.2 and z2.3 vs uninfected](figures/nodrug_upset.svg)

## Figure 3A Volcano z2.3+SbV / z2.3-SbV

FIXME: Olga noted that the ggbreak showing BOLA2B is not quite properly
placed (perhaps also 3C)

Section 12.5.7

I think I reversed the numerator/denominator when plotting this with
labels.  I guess it does not particularly matter as long as the genes
are at the correct logFC/p-value, but it is an odd discrepancy.  Ah,
it looks like my latest revision changed it back.

Here is the relevant code block:

```{r, eval=FALSE}
labeled <- z23sb_vs_z23nosb_volcano[["plot"]] +
  scale_x_continuous(limits = c(-19, 6),
                     breaks = c(-20, -18, -16, -14, -12, -10, -6, -4, -2, 0, 2, 4, 6)) +
  ggbreak::scale_x_break(c(-17, -8), scales = 17, space = 0.02)
pp(file = "figures/fig3a_labeled_with_break.pdf")
labeled
dev.off()
```

and plot:

![Volcano comparing z2.3+SbV and z2.3-SbV](figures/fig3a_labeled_with_break.svg)

## Figure 3B Volcano z2.2+SbV / z2.2-SbV

This is the analagous plot but the zymodeme 2.2 infections, here is
the relevant code block:

Oddly, I did not see this get exported to a svg/pdf?

```{r, eval=FALSE}
z22sb_vs_z22nosb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z22sb_vs_z22nosb"]], "z22sb_vs_z22nosb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol",
  color_high = plot_colors[["infsbz22"]], color_low = plot_colors[["infz22"]])

labeled <- z22sb_vs_z22nosb_volcano[["plot"]] +
  scale_x_continuous(breaks = c(-6, -4, -2, 0, 2, 4, 6))

pp(file = "figures/fig3b_labeled.svg")
labeled
dev.off()
labeled
```

![Volcano comparing z2.2+SbV and z2.2-SbV](figures/fig3b_labeled.svg)

## Figure 3C Volcano of z2.3+SbV / z2.2+SbV

FIXME: Olga noted that the ggbreak showing BOLA2B is not quite properly
placed (perhaps also 3A)

If you wish to find the relevant code:

```{r, eval=FALSE}
z23sb_vs_z22sb_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["z23sb_vs_z22sb"]], "z23sb_vs_z22sb",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol",
  color_high = plot_colors[["infsbz23"]], color_low = plot_colors[["infsbz22"]])

labeled <- z23sb_vs_z22sb_volcano[["plot"]] +
  scale_x_continuous(breaks = c(-23, -6, -4, -2, 0, 2, 4, 6)) +
  ggbreak::scale_x_break(c(-5, -22.5), scales = 10, space = 0.02)
pp(file = "figures/fig3c_labeled_breaks.svg")
labeled
dev.off()
labeled
```

![Volcano comparing z2.3+SbV and z2.2+SbV](figures/fig3c_labeled_with_breaks.svg)

## Figure 2D Volcano of uninfected+SbV / uninfected-SbV

The relevant code block, which afaict was not previously made
explicit:

```{r, eval=FALSE}
sb_vs_uninf_volcano <- plot_volcano_condition_de(
  hs_macr_table[["data"]][["sb_vs_uninf"]], "sb_vs_uninf",
  fc_col = "deseq_logfc", p_col = "deseq_adjp",
  label = 10, label_column = "hgncsymbol",
  color_high = plot_colors[["sb"]], color_low = plot_colors[["uninf"]])

labeled <- sb_vs_uninf_volcano[["plot"]] +
  scale_x_continuous(breaks = c(-23, -6, -4, -2, 0, 2, 4, 6)) +
  ggbreak::scale_x_break(c(-5, -22.5), scales = 10, space = 0.02)
pp(file = "figures/fig3d_labeled_breaks.svg")
labeled
dev.off()
labeled
```

The resulting image:

![Volcano comparing Uninf+SbV and Uninf-SbV](figures/fig3d_labeled_breaks.svg)

## Figure 3E UpSet showing unique/shared genes A-D

The relevant code portion:

```{r, eval=FALSE}
drugnodrug_uninf_contrasts <- c("z23sb_vs_z23nosb", "z22sb_vs_z22nosb", "sb_vs_uninf")
drugnodrug_upset_with_uninf <- upsetr_combined_de(hs_macr_table,
                                       desired_contrasts = drugnodrug_uninf_contrasts)
pp(file = "figures/drugnodrug_with_uninf_upset.svg")
drugnodrug_upset_with_uninf
dev.off()
drugnodrug_upset_with_uninf
```

The resulting image:

![UpSet comparing drug/nodrug including uninfected](figures/drugnodrug_with_uninf_upset.svg)

## Figure 4

I think these were produced by Dr. Colmenares.

## Table S1: Metadata

This may be found here:

[Sample Metadata](sample_sheets/tmrc2_msamples_pruned.xlsx "Sample Metadata.")

## Table S2: DE Tables

This is produced in 03differential_expression.Rmd in the block which
looks like:

```{r, eval=FALSE}
#test = deseq_pairwise(normalize_expt(hs_macr, filter=TRUE),
#                      model_batch = "svaseq", filter = TRUE,
#                      extra_contrasts = tmrc2_human_extra)

hs_macr_de <- all_pairwise(hs_macr, model_batch = "svaseq",
                           filter = TRUE, extra_contrasts = tmrc2_human_extra)
hs_macr_de
hs_single_table <- combine_de_tables(
  hs_macr_de, keepers = single_tmrc2_keeper,
  excel = glue("analyses/macrophage_de/{ver}/de_tables/hs_macr_drug_zymo_z22sb_sb-v{ver}.xlsx"))
hs_single_table
hs_macr_table <- combine_de_tables(
  hs_macr_de, keepers = tmrc2_human_keepers,
  excel = glue("analyses/macrophage_de/de_tables/hs_macr_drug_zymo_table_macr_only.xlsx"))
hs_macr_table
```

[Differential Expression Results](analyses/macrophage_de/de_tables/hs_macr_drug_zymo_table_macr_only.xlsx "Differential Expression Results.")

## Table S3

This is a curated set of ontologies, I think it is coming from a mix of Dr. Colmenares and my own GSEA/enrichment analyses.

# Bibliography
